Data Migration under a Changing Schema in Ampersand

### 2023-11-28 2022Migration/articleMigrationFoIKS.tex
in essence submittable

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\maintain}{\mathbin{\id{maint}}}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\viol}[2]{\violC{#1}(#2)}
\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}_{#1}}
\newcommand{\enfRel}[1]{\id{enf}_{#1}}
\newcommand{\signature}[2]{\langle{#1},{#2}\rangle}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quintuple}[5]{\langle{#1},{#2},{#3},{#4},{#5}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\transactions}{\mathcal{E}}
\newcommand{\busConstraints}{\mathcal{B}}
\newcommand{\Constraints}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   \begin{figure}[bht]
      \begin{center}
        \includegraphics[scale=0.8]{figures/existing system.pdf}
      \end{center}
   \caption{Anatomy of an information system}
   \label{fig:pre-migration}
   \end{figure}

@@

   We assume that constraints implement the business semantics of the data.
   Constraints represent business concerns formally, so they can be checked automatically and can be used to generate software.
   Some of the constraints require human intervention, while others require a system to intervene.
   In this paper, we distinguish three different kinds of constraints:
\begin{enumerate}
\item Blocking invariant\\
   A \define{blocking invariant} is a constraint that is always true in a system.
   It serves to constrain the data set at runtime.
   When the data set changes in a way that violates a blocking invariant, the system produces an error message and refuses the change.
\item Transactional invariant\\
   The classical database transaction can be understood as a \define{transactional invariant},
   which the system keeps satisfied by adding or deleting triples to the dataset.
   As soon as the data violates this constraint, the system restores it without human intervention.
   So, the outside world experiences this as a constraint that is always true, i.e.~an invariant.
\item Business constraint\\
   A \define{business constraint} is a constraint that users can violate temporarily until someone restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").
   That takes some time, during which the constraint is violated.
   So, we do not consider business constraints to be invariants.
\end{enumerate}
   Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
   Of the three types of constraint, only two are invariants.

@@

   Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
   It encompasses the necessary constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
   The systems they generate maintain invariants and alert users to violations of business constraints.
   Ampersand's explicit representation of all invariants in its source code makes it well-suited for implementing our theory.
   The absence of imperative code in Ampersand scripts enhances system reasoning,
   while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
   Constraints, which carry business semantics, make ``preserving the meaning as much as possible'' explicit.
   An Ampersand script provides just enough information to generate a complete system,
   allowing extraction of a classical database schema (i.e., data structure plus constraints) from the Ampersand script.

@@

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero downtime, to enable more frequent SCDMs.
\item users must be able to fix invariants in the new system.
   So, we need an intermediate ``migration system'' that implements blocking constraints of the desired system as business constraints,
   in order to deliver zero downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time,
   to ensure that the migration system does not need to live ever after.
\end{enumerate}

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations typically deploy the existing system and the desired system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.8]{figures/migration system deployed.pdf}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible and to achieve zero downtime,
   we must deploy a third system: the migration system.
   This system has its own schema.
   It comprises the schemas of both the existing system and the desired system,
   so the migration system can copy the data from the existing to the desired system.
   It uses transactions to copy the data and to resolve some forms of data pollution.
   Not all migration work, however, can be automated.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require tailoring the migration script to specific needs
   that require action of users in production.
   For that purpose, the migration engineer specifies business constraints in the migration system.
   In Ampersand, a developer can use business constraints to implement such needs.

@@

\section{Definitions}
\label{sct:Definitions}
   An \define{information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, we ignore functionality captured in user interfaces because it does not impact the migration.
   Section~\ref{sct:Data sets} describes how we define data sets.
   Since data sets are sets, we will use set operators $\cup$ (union), $\cap$ (intersect), $-$ (set difference), and an overline $\cmpl{x}$ as complement operator on sets.
   Section~\ref{sct:Constraints} defines constraints and their violations.
   Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

\subsection{Constraints}
\label{sct:Constraints}
   Every constraint is an element of an infinite set called $\Constraints$.
   In this paper, variables $u$ and $v$ represent all three types of constraints.
   We say that a constraint, $u$, is satisfied when $u$ is true for all triples in the data set.
   For every constraint $u$, function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produces the violations of $u$,
   and $\sign{u}:\Pair{\Concepts}{\Concepts}$ yields the signature of $u$.
   The definition of $\violC{u}$ implies the assumption that we represent each violation as a pair.
   Every constraint must satisfy:
\begin{equation}
   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\label{eqn:wellTypedViolation}
\end{equation}
   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.

   In case $u$ is a transactional invariant,
   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$.
\begin{equation}
   \begin{array}[t]{l}
      \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
      \text{\bf with}\ \instance'=\{ \pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}
   \end{array}
\label{eqn:transaction}
\end{equation}
   It is obvious that not every conceivable constraint can satisfy this equation.
   So, we assume that the compiler restricts the set of transactional invariants to those that satisfy equation~\ref{eqn:transaction}.
   As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
   We call this the \define{enforced relation} of transactional invariant $u$:
\begin{equation}
   \enfRel{u}=\declare{n}{A}{B}
\end{equation}
   Let us denote a transactional invariant as $r\mapsfrom\violC{u}$ or equivalently $r \mapsfrom \lambda \dataset.~\viol{u}{\dataset}$,
   in which $r=\enfRel{u}$.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item Transactional invariants remain satisfied by adding violations to a specific relation (\ref{eqn:transaction}).
\item All invariants must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\end{itemize}
\end{definition}
   We assume that a deployment will fail if these requirements are not satisfied.
   We will write $\text{isIS}(\infsys)$ to denote that $\infsys$ is an information system.

\section{Generating a Migration Script}
\label{sct:Generating}
   The complexity of migrating data to date yields expensive and error-prone migration projects.
   By generating this code we can eliminate human induced errors.

@@

\subsection{Algorithm for generating a migration script}
   In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
   We have to relabel items with prefixes to avoid name clashes in the migration system.
   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.
\begin{equation}
   \begin{array}[m]{rcl}
      \overleftarrow{\pair{\dataset}{\schema}}&=&\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}\\
      \overleftarrow{\pair{\triples}{\instance}}&=&\pair{\overleftarrow{\triples}}{\instance}\\
      \overleftarrow{\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}}&=&\quintuple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}{\overleftarrow{\busConstraints}}\\
      \overleftarrow{\triples}&=&\{\triple{a}{\overleftarrow{r}}{b}\mid\triple{a}{r}{b}\in\triples\}\\
      \overleftarrow{\declare{n}{A}{B}}&=&\declare{old.n}{A}{B}\\
      \overleftarrow{X}&=&\{\overleftarrow{x}\mid x\in X\}\\
      \viol{\overleftarrow{u}}{\overleftarrow{\dataset}}&=&\viol{u}{\dataset}\\
      \sign{\overleftarrow{u}}&=&\sign{u}\\
      \enfRel{\overleftarrow{u}}&=&\overleftarrow{\enfRel{u}}
   \end{array}
\end{equation}
   The notation $\overrightarrow{\infsys}$ is defined similarly, using prefix ``new.'' instead of ``old.''.

>>>>

%\def\id#1{\text{\it #1\/}}
\newcommand{\xrightarrowdbl}[2][]{%
  \xrightarrow[#1]{#2}\mathrel{\mkern-14mu}\rightarrow
}
\newcommand{\id}[1]{\text{\it #1\/}}
\newcommand{\code}[1]{\text{\tt\small #1}}
\newcommand{\stmtText}[1]{``{\small\tt #1}''}
\newcommand{\dom}[1]{\id{dom}(#1)}
\newcommand{\cod}[1]{\id{cod}(#1)}
%\renewcommand{\int}[2]{\id{inter}(#1,#2)}
\newcommand{\relsIn}[1]{\id{relsIn}(#1)}    % maps a Term to a set of Relations
\newcommand{\popF}[1]{\id{pop}_{#1}}
\newcommand{\pop}[2]{\popF{#1}(#2)}
\newcommand{\enf}{\mathbin{{\tt enforce}}}
\newcommand{\enforce}[2]{{\tt enforce}_{#1}(#2)}
\newcommand{\enforceC}[1]{{\tt enforce}_{#1}}
\newcommand{\instance}{\mathbin{\id{inst}}}
\newcommand{\relname}[1]{\id{relname}(#1)}
\newcommand{\evt}[2]{\id{event}_{#1,#2}}
\newcommand{\src}[1]{\id{src}(#1)}
\newcommand{\tgt}[1]{\id{tgt}(#1)}
\newcommand{\viol}[2]{\violC{#1}(#2)}
\newcommand{\violC}[1]{\id{viol}_{#1}}
\newcommand{\sign}[1]{\id{sign}_{#1}}
\newcommand{\enfRel}[1]{\id{enf}_{#1}}
\newcommand{\signature}[2]{\langle{#1},{#2}\rangle}
\newcommand{\powerset}[1]{\cal{P}\{#1\}}
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\full}{V}
\newcommand{\declare}[3]{\id{#1}_{\pair{#2}{#3}}}
\newcommand{\subst}[3]{#3_{[#1\rightarrow #2]}}
\newcommand{\fullt}[2]{V_{\pair{#1}{#2}}}
\newcommand{\iden}{I}
\newcommand{\ident}[1]{I_{\id{#1}}}
\newcommand{\expr}[3]{(#1)_{#2\times #3}}
\newcommand{\pair}[2]{\langle{#1},{#2}\rangle}
\newcommand{\maprel}[2]{{\tt maprel}_{#1}({#2})}
\newcommand{\Pair}[2]{#1\times#2}
\newcommand{\pairs}[1]{\id{pairs}(#1)}
\newcommand{\triple}[3]{\langle{#1},{#2},{#3}\rangle}
\newcommand{\quintuple}[5]{\langle{#1},{#2},{#3},{#4},{#5}\rangle}
\newcommand{\atom}[1]{{\tt\small #1}}
\newcommand{\atoms}{\mathcal{A}}
\newcommand{\Atoms}{\mathbb{A}}
%\newcommand{\events}{\mathcal{E}}
%\newcommand{\Events}{\mathbb{E}}
\newcommand{\concept}[1]{{\tt\small #1}}
\newcommand{\concepts}{\mathcal{C}}
\newcommand{\Concepts}{\mathbb{C}}
\newcommand{\decls}{\mathcal{D}}  %% names of relations
\newcommand{\rels}{\mathcal{R}}   %% all relations
\newcommand{\Rels}{\mathbb{R}}   %% all relations
\newcommand{\relations}{\mathcal{M}} % representing terms. M is a subset of R.
\newcommand{\triples}{\mathcal{T}}
\newcommand{\Triples}{\mathbb{T}}
\newcommand{\Triple}[3]{#1\times#2\times#3}
\newcommand{\vertices}{N}
\newcommand{\rules}{\mathcal{U}}
\newcommand{\transactions}{\mathcal{E}}
\newcommand{\busConstraints}{\mathcal{B}}
\newcommand{\Constraints}{\mathbb{U}}
\newcommand{\specrules}{\mathcal{S}}
\newcommand{\roles}{\mathcal{O}}
\newcommand{\dataset}{\mathscr{D}}
\newcommand{\Dataset}{\mathbb{D}}
\newcommand{\schema}{\mathscr{Z}}
\newcommand{\functionality}{\mathscr{F}}
\newcommand{\select}[2]{\id{select}_{#1}\{{#2}\}}
\newcommand{\migrsys}{\mathscr{M}}
\newcommand{\infsys}{\mathscr{S}}
\newcommand{\tf}[1]{\mathscr{T}(#1)}
\newcommand{\ptf}[1]{\mathscr{T}'(#1)}
\newcommand{\ti}[1]{\mathscr{I}(#1)}
\newcommand{\tic}[1]{I_{\cal C}(#1)}
\newcommand{\relAdd}{\dagger}
\newcommand{\flip}[1]{{#1}^\smallsmile} %formerly:  {#1}^\backsim
\newcommand{\kleeneplus}[1]{{#1}^+}
\newcommand{\kleenestar}[1]{{#1}^*}
\newcommand{\cmpl}[1]{\overline{#1}}
\newcommand{\rel}{\times}
\newcommand{\compose}{;}
\newcommand{\subs}{\subseteq}%{\models}
\newcommand{\fun}{\rightarrow}
\newcommand{\isa}{\preceq}
%\newcommand{\isaClos}{\sqsubseteq}
\newcommand{\typetest}{?}
\newcommand{\meet}{\sqcap}
\newcommand{\join}{\sqcup}
\newcommand{\Meet}{\bigsqcap}
\newcommand{\Moin}{\bigsqcup} % because LaTeX has already defined command \Join.
\newcommand{\order}{\ominus}
\newcommand{\anything}{\top}
\newcommand{\nothing}{\bot}
\newcommand{\rewriteto}{\rightarrow}
\newcommand{\calc}{\implies}
\newcommand{\alland}{\bigwedge}
\newcommand{\mph}[3]{#1_{#2\times #3}}
\newcommand{\mphu}[1]{#1_{\univ\times\univ}}

@@

\section{Analysis}
\label{sct:Analysis}
   This section analyzes information systems qualitatively, to prepare for a formal treatment in section~\ref{sct:Definitions}.
   The current section yields a procedure for migrating data from one system to another.
\subsection{Information Systems}
   The purpose of an information system is to store and disclose data in a way that is meaningful to its users.
   Multiple users, working from different locations and on different moments, constitute what we will loosely call ``the business''.
   The data in the system constitutes the collective memory of the business,
   which relies on the semantics of the data to draw the right conclusions and carry out their tasks.
   \begin{figure}[bht]
      \begin{center}
        \includegraphics[scale=0.8]{figures/existing_system.pdf}
      \end{center}
   \caption{Anatomy of an information system}
   \label{fig:pre-migration}
   \end{figure}

@@

   We assume that constraints implement the business semantics of the data.
   Constraints represent business concerns formally, so they can be checked automatically and can be used to generate software.
   Some of the constraints require human intervention, while others require a system to intervene.
   In this paper, we distinguish three different kinds of constraints:
\begin{enumerate}
\item Blocking invariant\\
   A \define{blocking invariant} is a constraint that is always satisfied in a system.
   It serves to constrain the data set at runtime.
   When the data set changes in a way that violates a blocking invariant, the system produces an error message and refuses the change.
\item Transactional invariant\\
   The classical database transaction can be understood as a \define{transactional invariant},
   which the system keeps satisfied by adding or deleting triples to the dataset.
   As soon as the data violates this constraint, the system restores it without human intervention.
   So, the outside world experiences this as a constraint that is always satisfied, i.e.~an invariant.
\item Business constraint\\
   A \define{business constraint} is a constraint that users can violate temporarily until someone restores it.
   Example: ``An authorized manager has to sign every purchase order.''
   Every violation requires some form of human action to satisfy the business constraint (e.g. "sign the purchase order").
   That takes some time, during which the constraint is violated.
   So, we do not consider business constraints to be invariants.
\end{enumerate}
   Summarizing, in our notion of information systems, concepts, relations, and constraints carry the business semantics.
   Of the three types of constraint, only two are invariants.

@@

   Ampersand serves as a language for specifying information systems through a framework of concepts, relations, and constraints.
   It comprises the three types of constraints discussed in this paper, making it an ideal platform for practical testing of our theory.
   In Ampersand, developers articulate constraints using heterogeneous relation algebra~\cite{Hattensperger1999,Alloy2006}
   The systems they generate keep invariants satisfied and alert users to violations of business constraints.
   The absence of imperative code in Ampersand scripts enhances reasoning about the system,
   while its static typing~\cite{vdWoude2011} yields the established benefits in software engineering processes~\cite{HanenbergKRTS14,Petersen2014}.
   Constraints carry business semantics, which makes ``preserving the meaning as much as possible'' explicit.
   An Ampersand script provides just enough information to generate a complete system,
   allowing extraction of a classical database schema (i.e., data structure plus constraints) from the Ampersand script.

@@

   Summarizing, the following requirements apply to SCDMs
\begin{enumerate}
\item users must experience zero downtime, to enable more frequent SCDMs.
\item users must be able to restore invariants in the new system.
   So, we need an intermediate ``migration system'' that implements blocking constraints of the desired system as business constraints,
   in order to deliver zero downtime.
\item the number of violations that users must fix is finite and decreases (monotonically) over time,
   to ensure that the migration system does not need to be kept alive infinitely.
\end{enumerate}

\subsection{Data Migrations}
   Data migration occurs when a desired system replaces an existing one,
   while preserving the meaning of the present data as much as possible~\cite{Spivak2012}.
   %Just copying the set of data from the existing system to the desired system is obviously wrong if the schemas of both systems differ.
   %
   In practice, data migrations typically deploy the existing system and the desired system side-by-side,
   while transferring data in a controlled fashion, as shown in Figure~\ref{fig:migration phase}.
\begin{figure}[bht]
   \begin{center}
     \includegraphics[scale=.8]{figures/migration_system_deployed.pdf}
   \end{center}
\caption{Migration phase}
\label{fig:migration phase}
\end{figure}
   To automate the migration as much as possible and to achieve zero downtime,
   we must deploy a third system: the migration system.
   This system has its own schema.
   It comprises the schemas of both the existing system and the desired system,
   so the migration system can copy the data from the existing to the desired system.
   It uses transactions to copy the data and to resolve some forms of data pollution.
   Not all of the work, however, can be automated.
   Data pollution, new business rules, or known issues in the existing system
   may occasionally require tailoring the migration script to specific needs
   that require action of users in production.
   For that purpose, the migration engineer specifies business constraints in the migration system.
   In Ampersand, a developer can use business constraints to implement such needs.

@@

\section{Definitions}
\label{sct:Definitions}
   An \define{information system} is a combination of data set, schema, and functionality.
   For the purpose of this paper, we ignore functionality captured in user interfaces to focus on the transition of business semantics in the data.
   Section~\ref{sct:Data sets} describes how we define data sets.
   Since data sets are sets, we will use set operators $\cup$ (union), $\cap$ (intersect), $-$ (set difference), and an overline $\cmpl{x}$ as complement operator on sets.
   Section~\ref{sct:Constraints} defines constraints and their violations.
   Schemas are treated in section~\ref{sct:Schemas}.
   Then section~\ref{sct:Information Systems} defines information systems.

@@

\subsection{Constraints}
\label{sct:Constraints}
   Every constraint is an element of an infinite set called $\Constraints$.
   In this paper, variables $u$ and $v$ represent all three types of constraints.
   We say that a constraint, $u$, is satisfied when it is true for all triples in the data set.
   For every constraint $u$, function $\violC{u}:\Dataset\rightarrow\powerset{\Pair{\Atoms}{\Atoms}}$ produces the violations of $u$,
   and $\sign{u}:\Pair{\Concepts}{\Concepts}$ yields the signature of $u$.
   The definition of $\violC{u}$ implies the assumption that we represent each violation as a pair.
   Every constraint must satisfy:
\begin{equation}
   \pair{a}{b}\in\viol{u}{\dataset}\ \wedge\ \sign{u}=\pair{A}{B}\ \Rightarrow\ a\instance A\ \wedge\ b\instance B
\label{eqn:wellTypedViolation}
\end{equation}
   Note that $\viol{u}{\dataset}=\emptyset$ means that $\dataset$ satisfies $u$.

   In case $u$ is a transactional invariant,
   the system will keep it satisfied by adding the violations to a specific relation $\declare{n}{A}{B}$.
\begin{equation}
   \begin{array}[t]{l}
      \viol{u}{\pair{\triples\cup\{\triple{a}{\declare{n}{A}{B}}{b}\mid\pair{a}{b}\in\viol{u}{\dataset}\}}{\instance\cup\instance'}}=\emptyset\\
      \text{\bf with\ }\begin{array}[t]{lcl}
         \instance&=&\{\pair{a}{A}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\cup\{ \pair{b}{B}\mid\pair{a}{b}\in\viol{u}{\dataset}\}\\
         \pair{A}{B}&=&\sign{u}
      \end{array}
   \end{array}
\label{eqn:transaction}
\end{equation}
   It is obvious that not every conceivable constraint can satisfy this equation.
   So, we assume that the compiler restricts the set of transactional invariants to those that satisfy equation~\ref{eqn:transaction}.
   As $\declare{n}{A}{B}$ is specific for $u$, we can write $\enfRel{u}$ for it.
   We call this the \define{enforced relation} of transactional invariant $u$:
\begin{equation}
   \enfRel{u}=\declare{n}{A}{B}
\end{equation}
   Let us denote a transactional invariant as $r\mapsfrom\violC{u}$ or equivalently $r \mapsfrom \lambda \dataset.~\viol{u}{\dataset}$,
   in which $r=\enfRel{u}$.

@@

\subsection{Information Systems}
\label{sct:Information Systems}
   Let us now define information systems by enumerating the requirements.
\begin{definition}[information system]
\label{def:information system}
\item An information system $\infsys$ is a tuple $\pair{\dataset}{\schema}$, in which
\begin{itemize}
\item $\dataset=\pair{\triples}{\instance}$ is a data set (so it must satisfy equation~\ref{eqn:wellTypedEdge}).
   We write $\triples_\infsys = \triples$ and $\instance_\infsys = \instance$ if needed;
\item $\schema=\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}$ is a schema (so it must satisfy equations~\ref{eqn:relationsIntroduceConcepts} thru~\ref{eqn:enforcement-has-type}).
\item Triples in the data set must have their relation mentioned in the schema:
   \begin{eqnarray}
      \triple{a}{\declare{n}{A}{B}}{b}\in\triples&\Rightarrow&\declare{n}{A}{B}\in\rels
   \label{eqn:define R}
   \end{eqnarray}
\item All violations must have a type, which follows from  (\ref{eqn:wellTypedViolation}).
\item The system keeps any transactional invariant $u$ satisfied by adding violations to the relation $\enforceC{u}$ (\ref{eqn:transaction}).
\item All invariants must remain satisfied:
   \begin{align}
      \forall u\in\rules\cup\transactions&.~\viol{u}{\dataset}=\emptyset
   \label{eqn:satisfaction}
   \end{align}
\end{itemize}
\end{definition}
   We assume that a deployment will fail if these requirements are not satisfied.

\section{Generating a Migration Script}
\label{sct:Generating}
   The complexity of migrating data to date yields expensive and error-prone migration projects.
   By generating this code we can prevent many human induced errors.

@@

\subsection{Algorithm for generating a migration script}
   In the migration system, we need to refer to the items (concepts, relations, and constraints) of both the existing system and the desired system.
   We have to relabel items with prefixes to avoid name clashes in the migration system.
   We use a left arrow to denote relabeling by prefixing the name of the item with ``old.''.
\begin{equation}
   \begin{array}[m]{rcl}
      \overleftarrow{\pair{\dataset}{\schema}}&=&\pair{\overleftarrow{\dataset}}{\overleftarrow{\schema}}\\
      \overleftarrow{\pair{\triples}{\instance}}&=&\pair{\overleftarrow{\triples}}{\instance}\\
      \overleftarrow{\quintuple{\concepts}{\rels}{\rules}{\transactions}{\busConstraints}}&=&\quintuple{\concepts}{\overleftarrow{\rels}}{\overleftarrow{\rules}}{\overleftarrow{\transactions}}{\overleftarrow{\busConstraints}}\\
      \overleftarrow{\triples}&=&\{\triple{a}{\overleftarrow{r}}{b}\mid\triple{a}{r}{b}\in\triples\}\\
      \overleftarrow{\declare{n}{A}{B}}&=&\declare{old.n}{A}{B}\\
      \overleftarrow{X}&=&\{\overleftarrow{x}\mid x\in X\}\\
      \viol{\overleftarrow{u}}{\overleftarrow{\dataset}}&=&\viol{u}{\dataset}\\
      \sign{\overleftarrow{u}}&=&\sign{u}\\
      \enfRel{\overleftarrow{u}}&=&\overleftarrow{\enfRel{u}}
   \end{array}
\end{equation}
   The notation $\overrightarrow{\infsys}$ is defined similarly, using prefix ``new.'' instead of ``old.''.
   This relabeling does not have any effect on the behavior of the system, i.e. $\infsys$, $\overleftarrow{\infsys}$, and $\overrightarrow{\infsys}$ are indistinguishable for end-users.
